---
title: "Reproducible Research: Peer Assessment 2"
output: html_document
keep_md: true
---
## Synopsis
This report concludes findings about severe weather events and their impact on public health and economic consequences. The findings are based on the U.S. National Oceaninc and Atmospheric Administration's (NOAA) storm database tracking characteristics of major storms and weather events in the United States. Key information includes place and time of observed events as well as estimates of any fatalities, injuries, crop and property damage.

## Configuration
```{r}
# Show the code
echo = TRUE
```

## Utility Functions
```{r}
# Quantifies multiplication symbols as an nth power of ten.
multiply <- function(m, x) {
  m_val <- switch(m, 
                  "h" = 2,
                  "k" = 3,
                  "m" = 6,
                  "b" = 9,
                  0)
  return (x * 10^m_val)
}

# Draws a bar chart.
# df        .. data frame containing the plotted data
# weight    .. columnar weight
# ylab      .. y-axis label
# ylab_unit .. y-axis unit (optional)
# title     .. a distinct part of a chart title
bar_chart <- function(df, weight, ylab, ylab_unit = NA, title, filename) {
  ylab_title <- paste(ylab, title)
  if (!is.na(ylab_unit)) {
    ylab_title <- paste(ylab_title, "in", ylab_unit)
  }
  qplot(EVTYPE, data = df, weight = weight, geom = "bar", binwidth = 1) + 
           scale_y_continuous(ylab_title) + 
           theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
           xlab("Extreme Weather Event") + 
           ggtitle(paste(title, 
                         "by Extreme Weather Events in the US\nbetween", 
                         year_from, "and", year_to))
}

# Transforms a column name into a description
as_title <- function(x) {
  # Capitalize
  title <- strsplit(x, " ")[[1]]
  title <- paste(toupper(substring(title, 1,1)), tolower(substring(title, 2)), sep = "")
  
  # Handle special parts
  gsub("dmg$", " Damage", title)
}

# Summarizes weather events by the given health or economic aspect. 
# Results are plotted as a histogram.
sum_by_aspect <- function(x, ylab, ylab_unit = NA, cap = 10, debug = FALSE) {
  # Summarize weather events by the given health aspect
  summary <- aggregate(subset(storm_data, select = x), list(EVTYPE = storm_data$EVTYPE), "sum")
  
  # Sort in a descending order (most frequent first)
  setorderv(summary, x, -1)
  
  # Consider Top N (Top 10 by default) most severe events
  top_n <- head(summary, cap)
  
  if (debug) {
    print(top_n)  
  }
  
  # Plot it on a bar chart
  bar_chart(top_n, top_n[, x], ylab, ylab_unit, as_title(x), tolower(x))
}
```

## Dependencies
```{r}
# Verifies that the required libraries get installed if needed
libs <- function(...) {
  lapply(list(...), function(lib) {
    if (!lib %in% installed.packages()) {
      install.packages(lib)
    }
    library(lib, character.only = TRUE)
  })
}

## Import all libraries
libs(
  "R.utils",                                 # To decompress the downloaded archive
  "data.table",                              # Data manipulation
  "dplyr", "ggplot2", "grid", "gridExtra"    # Plotting
)
```

## Data Processing
Download and decompress source data, unless that's been already done.
```{r}
storm_data_file <- "StormData.csv"
if (!file.exists(storm_data_file)) {
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  dest_file <- "StormData.csv.bz2"
  download.file(url, destfile = dest_file, method = "curl")
  bunzip2(dest_file, overwrite = TRUE, remove = TRUE)
}
```

Load the data (CSV file) into memory.

```{r}
storm_data <- fread(storm_data_file, header = TRUE, sep = ",", verbose = TRUE)
```

Add a new column representing a year. That will help to find out about most extreme periods in terms of sever weather conditions.

```{r}
storm_data$YEAR <- as.numeric(
  format(as.Date(storm_data$BGN_DATE, "%m/%d/%Y %H:%M:%S"), "%Y"))
```

To speed up processing time and reduce memory footprint table columns are stripped down to a bare minimum, i.e. only those needed to answer questions about extreme weather events.

```{r}
health_data <- c("FATALITIES", "INJURIES")          # Health
economic_data <- c("PROPDMG", "CROPDMG")            # Economy
economic_data_mult <- c("PROPDMGEXP", "CROPDMGEXP") # Economy (extent)

storm_data <- subset(storm_data, select=c(
  "YEAR",                         # Timeline
  "EVTYPE",                       # Weather Events 
  health_data,
  economic_data,
  economic_data_mult      
))
```

PROPDMGEXP and CROPDMGEXP represent the scale of property and crop damage. They are effectively multipliers of the values in PROPDMG and CROPDMG columns.

In order to quantify the economic impact the multipliers need to be standardized.

```{r}
storm_data <- mutate_each(storm_data, funs(tolower), 
                          which(colnames(storm_data) %in% economic_data_mult))
```

Now we are in a position to quantify the damage by using a custom utility function called 'multiply' (see the section: Utility Functions).

```{r}
storm_data <- mutate(storm_data, PROPDMG = mapply(multiply, storm_data$PROPDMGEXP, storm_data$PROPDMG))
storm_data <- mutate(storm_data, CROPDMG = mapply(multiply, storm_data$CROPDMGEXP, storm_data$CROPDMG))
```

Let's plot a histogram to easily identify the years in scope of the analysis.

```{r}
hist(storm_data$YEAR, 
     breaks = 20, 
     main = "Extreme Weather Events per Year",
     xlab = "Year")
```

The most extreme weather events happened between 1990 - 2010. Let's limit the observations to the found time span.

```{r}
year_from <- 1990
year_to <- 2010
storm_data <- subset(storm_data, (YEAR >= year_from) & (YEAR <= year_to))
```

Now we are ready to answer the question about the most harmful events with respect to population health. The following code will capture top ten events in these regards and plot them as a single figure.

```{r}
lapply(health_data, function(x) {
  sum_by_aspect(x, "Number of")
})
```

Finally, we look at the events causing the most extensive economic damage. The code produces a single figure providing an overview of severe weather conditions having a considerable economic impact.

```{r}
lapply(economic_data, function(x) {
  sum_by_aspect(x, "Cost of", "US Dollars")
})
```

## Results
TODO - fix the figures first!