---
title: "Impact of Severe Weather Events on U.S. Health and Economy"
author: Tomas Zezula, November 2015
output: 
  html_document:
    toc: true
keep_md: true
---
## Synopsis
This report concludes findings about severe weather events and their impact on public health and economic consequences. The findings are based on the U.S. National Oceaninc and Atmospheric Administration's (NOAA) storm database tracking characteristics of major storms and weather events in the United States. Key information includes place and time of observed events as well as estimates of any fatalities, injuries, crop and property damage.

## Configuration
```{r}
# Trace code execution
echo = TRUE

# Overrides scientific format on the y-axis in plots
options("scipen" = 20)

# Number formatting, limits the number of decimal places
options("digits" = 3)
```

## Utility Functions
```{r}
# Quantifies multiplication symbols in billions (of US Dollars).
multiply <- function(m, x) {
  m_val <- switch(m, 
                  "h" = 2,
                  "k" = 3,
                  "m" = 6,
                  "b" = 9,
                  0)
  return ((x * 10^m_val) / 10^9)
}

# Draws a bar chart.
# df        .. data frame containing the plotted data
# weight    .. columnar weight
# ylab      .. y-axis label
# ylab_unit .. y-axis unit (optional)
# title     .. a distinct part of a chart title
bar_chart <- function(df, weight, ylab, ylab_unit = NA, title) {
  ylab_title <- paste(ylab, title)
  if (!is.na(ylab_unit)) {
    ylab_title <- paste(ylab_title, "in", ylab_unit)
  }
  qplot(EVTYPE, data = df, weight = weight, geom = "bar", binwidth = 1) + 
         scale_y_continuous(ylab_title) + 
         theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
         xlab("Extreme Weather Event") + 
         ggtitle(paste(title, 
                       "\nby Extreme Weather Events in the US\nbetween", 
                       year_from, "and", year_to)) +
         theme(axis.title=element_text(size = 9), 
               plot.title = element_text(size = 11))
}

# Transforms a column name into a description
as_title <- function(x) {
  # Capitalize
  title <- strsplit(x, " ")[[1]]
  title <- paste(toupper(substring(title, 1,1)), tolower(substring(title, 2)), sep = "")
  
  # Handle special parts
  gsub("dmg$", " Damage", title)
}

# Summarizes weather events by the given health or economic aspect
# and returns results as a bar chart.
plot_by_aspect <- function(x, ylab, ylab_unit = NA, cap = 10, debug = TRUE) {
  # Summarize weather events by the given health aspect
  summary <- aggregate(subset(storm_data, select = x), list(EVTYPE = storm_data$EVTYPE), "sum")
  
  # Sort in a descending order (most frequent first)
  setorderv(summary, x, -1)
  
  # Consider Top N (Top 10 by default) most severe events
  top_n <- head(summary, cap)
  
  if (debug) {
    heading <- paste("###", as_title(x))
    if (!is.na(ylab_unit)) {
      heading <- paste(heading, "in", ylab_unit)
    }
    heading <- paste(heading, "###")
    print(heading)
    print(top_n)
  }
  
  # Plot it on a bar chart
  bar_chart(top_n, top_n[, x], ylab, ylab_unit, as_title(x))
}

# Plots a specific aspect of health impact of severe weather events.
# x .. healht aspect, such as injuries or fatalities
plot_health_impact <- function(x) {
  plot_by_aspect(x, "Number of")    
}

# Plots a specific aspect of economic impact of severe weather events.
# x .. economic aspect, such as crop or property damage
plot_damage <- function(x) {
  plot_by_aspect(x, "Cost of", "Billions of US Dollars")    
}

# Plots a group of related charts
# [varargs] .. one or more charts to be plotted as part ot the same figure
plot_group <- function(...) {
  grid.arrange(..., ncol = length(list(...)))
}
```

## Dependencies
```{r}
# Verifies that the required libraries get installed if needed
libs <- function(...) {
  lapply(list(...), function(lib) {
    if (!lib %in% installed.packages()) {
      install.packages(lib)
    }
    library(lib, character.only = TRUE)
  })
}

# Import all libraries
libs(
  "R.utils",                                 # To decompress the downloaded archive
  "data.table",                              # Data manipulation
  "dplyr", "ggplot2", "grid", "gridExtra"    # Plotting
)
```

## Data Processing
Download and decompress source data, unless that's been already done.
```{r}
storm_data_file <- "StormData.csv"
if (!file.exists(storm_data_file)) {
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  dest_file <- "StormData.csv.bz2"
  download.file(url, destfile = dest_file, method = "curl")
  bunzip2(dest_file, overwrite = TRUE, remove = TRUE)
}
```

Load the data (CSV file) into memory.

```{r}
storm_data <- fread(storm_data_file, header = TRUE, sep = ",", verbose = TRUE)
```

Add a new column representing a year. That will help to find out about most extreme periods in terms of sever weather conditions.

```{r}
storm_data$YEAR <- as.numeric(
  format(as.Date(storm_data$BGN_DATE, "%m/%d/%Y %H:%M:%S"), "%Y"))
```

To speed up processing time and reduce memory footprint table columns are stripped down to a bare minimum, i.e. only those needed to answer questions about extreme weather events.

```{r}
# Economic Multiplicators
economic_data_mult <- c("PROPDMGEXP", "CROPDMGEXP") 

# Reduce columns to only those needed for answering the questions
storm_data <- subset(storm_data, select=c(
  "YEAR",                         # Timeline
  "EVTYPE",                       # Weather Events 
  "FATALITIES", "INJURIES",       # Health
  "PROPDMG", "CROPDMG",           # Economy
  economic_data_mult))
```

PROPDMGEXP and CROPDMGEXP represent the scale of property and crop damage. They are effectively multipliers of the values in PROPDMG and CROPDMG columns.

In order to quantify the economic impact the multipliers need to be standardized.

```{r}
storm_data <- mutate_each(storm_data, funs(tolower), 
                          which(colnames(storm_data) %in% economic_data_mult))
```

## Results
Now we are in a position to quantify the damage by using a custom utility function called 'multiply' (see the section: Utility Functions).

```{r}
storm_data <- mutate(storm_data, PROPDMG = mapply(multiply, storm_data$PROPDMGEXP, storm_data$PROPDMG))
storm_data <- mutate(storm_data, CROPDMG = mapply(multiply, storm_data$CROPDMGEXP, storm_data$CROPDMG))
```

Let's plot a histogram to easily identify the years in scope of the analysis.

```{r}
hist(storm_data$YEAR, 
     breaks = 20, 
     main = "Extreme Weather Events per Year",
     xlab = "Year")
```

The most extreme weather events happened between 1990 - 2010. Let's limit the observations to the relevant time period.

```{r}
year_from <- 1990
year_to <- 2010
storm_data <- subset(storm_data, (YEAR >= year_from) & (YEAR <= year_to))
```

Now we are ready to answer the question about the most harmful events with respect to population health. The following code will capture top ten events in these regards and plot them as a single figure.

```{r}
plot_group(
  plot_health_impact("INJURIES"),
  plot_health_impact("FATALITIES")
)
```

Finally, we look at the events causing the most extensive economic damage. The code produces a single figure providing an overview of severe weather conditions having a considerable economic impact.

```{r}
plot_group(
  plot_damage("PROPDMG"),
  plot_damage("CROPDMG")
)
```

## Conclusion
The most severe weather conditions most harmful to **human health** were **tornados** and **excessive heat**. Tornados alone caused more than 20.000 injuries, whereas excessive heat proved to have been the most lethal factor leading to almost 2.000 fatalities between 1990 and 2010.

In terms of **economic consequences**, **flood** was the most devastating weather element causing close to 140 billion (USD) of **property damage**. Another dominant factor was **drought** causing an extensive **crop failure**.  

